---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To use `waspr` the user first needs to load the package as follows:

```{r message=FALSE}
library(waspr)
```

The user must provide a 3-dimensional array with posterior samples for all
parameters for each subset posterior (rows = subset posteriors, columns =
parameters, slices = samples). The amount of parameters and samples must be
equal for each subset posterior, these posterior samples may be obtained from
any type of MCMC algorithm. `waspr` provides an example array with posterior
samples for 8 parameters for 8 subset posteriors, `pois_logistic`, that will be
used to illustrate the functionality of the package.

The main function `wasp()` runs the swapping algorithm, combines its output and
computes the Wasserstein barycenter. It has four arguments, `mcmc`, that
specifies the 3-dimensional array with samples for each subset posterior, and
optional arguments `par.names`, that can be used to specify parameter names,
`iter` to specify the maximum number of iterations of the swapping algorithm,
`acc` to specify the accuracy of the swapping algorithm and `out` to indicate
whether the results per iteration of the swap algorithm should be printed.

```{r}

out <- wasp(pois_logistic,
            iter = 10,
            acc = 0.001,
            par.names = c("beta_s", "alpha_l", "beta_l",
                         "baseline_sigma", "baseline_mu",
                         "correlation", "sigma_s", "sigma_l"))

```

`wasp()` prints the iteration number and cost function value of the swapping
algorithm. The `out` object is of class `wasp` and contains several elements. To
obtain the Wasserstein barycenter of the subset posteriors a user can specify
`out$barycenter`. This returns a matrix of posterior samples (rows) for all
parameters (columns) of the full data posterior. A summary of the approximation
of the full data posterior is available through `summary(out)`.

```{r}
summary(out)
```
