---
title: "RelativePermeability"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette
bibliography: REFERENCES.bib
csl: apa-6th-edition.csl    
vignette: >
  %\VignetteIndexEntry{RelativePermeability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Rrelperm

`Rrelperm` generates relative permeability values for two- and/or three-phase mixtures of water and hydrocarbons. It estimates the two-phase relative permeabilities for the water-oil and gas-liquid systems in a tabular format using the modified Brooks-Corey correlation [@Brooks1964]. It calculates the three-phase relative permeabilities using the modified Stone I, Stone II, and Baker's linear models [@Stone1970; @Fayers1984; @Baker1988].<br>

## Relative Permeability Parameters

```{r, echo = FALSE, fig.width = 6, fig.height= 6, fig.align = "center", warning = FALSE, message = FALSE}
library(Rrelperm)
library(ggplot2)
library(magrittr)
library(tidyr)

rel_perm_ow <- kr2p_ow(SWCON = 0.13, SWCRIT = 0.20, SOIRW = 0.15, SORW = 0.22, 
                       KRWIRO = 0.45, KROCW = 1, NW = 1.5, NOW = 1.5, NP = 101)

rel_perm_ow_df <- as.data.frame(rel_perm_ow)
colnames(rel_perm_ow_df) <- c("Sw", "So", "Krw", "Krow")

p1 <- rel_perm_ow_df %>% 
  tidyr::pivot_longer(cols = c(Krw, Krow), values_to = "value",  names_to = "Kr") %>%
  ggplot(aes(x = Sw, y = value, col = Kr)) +
  geom_point() +
  labs(title = "water-oil relative permeability", x = "Sw", y = "Kr") +
  geom_segment(x = 0.0, y = 0.05, xend = 0.13, yend = 0.05, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.07, y=0.075, label="SWCON", size = 3, color="black") +
  geom_segment(x = 0.0, y = 0.1, xend = 0.2, yend = 0.1, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.105, y=0.125, label="SWCRIT", size = 3, color="black") +
  geom_segment(x = 0.85, y = 0.05, xend = 1.0, yend = 0.05, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.92, y=0.075, label="SOIRW", size = 3, color="black") +
  geom_segment(x = 0.78, y = 0.1, xend = 1.0, yend = 0.1, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.885, y=0.125, label="SORW", size = 3, color="black") +
  geom_segment(x = 0.65, y = 0.45, xend = 0.80, yend = 0.45, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "last", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.725, y=0.475, label = "KRWIRO", size = 3, color="black") +
  geom_segment(x = 0.18, y = 1.0, xend = 0.33, yend = 1.0, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "first", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.275, y = 0.975, label = "KROCW", size = 3, color="black") +
  theme_bw()

p1
```
 
 
 
```{r, echo = FALSE, fig.width = 6, fig.height= 6, fig.align = "center", warning = FALSE, message = FALSE}
library(Rrelperm)
library(ggplot2)
library(magrittr)

rel_perm_gl <- kr2p_gl(SWCON = 0.13, SOIRG = 0.10, SORG = 0.15, SGCON = 0.05, 
                       SGCRIT = 0.1, KRGCL = 0.45, KROGCG = 1.0, NG = 1.25, NOG = 1.25, 
                       NP = 101)

rel_perm_gl_df <- as.data.frame(rel_perm_gl)
colnames(rel_perm_gl_df) <- c("Sg", "Sl", "Krg", "Krog")

p2 <- rel_perm_gl_df %>% tidyr::pivot_longer(cols = c(Krg, Krog), values_to = "value", 
                                             names_to = "Kr") %>%
  ggplot(aes(x = Sg, y = value, col = Kr)) +
  geom_point() +
  labs(title = "gas-liquid relative permeability", x = "Sg", y = "Kr") +
  geom_segment(x = 0.0, y = 0.05, xend = 0.05, yend = 0.05, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.03, y=0.075, label="SGCON", size = 3, color="black") +
  geom_segment(x = 0.0, y = 0.1, xend = 0.1, yend = 0.1, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.05, y=0.125, label="SGCRIT", size = 3, color="black") +
  geom_segment(x = 0.77, y = 0.05, xend = 1.0, yend = 0.05, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.92, y=0.075, label="SLCON", size = 3, color="black") +
  geom_segment(x = 0.72, y = 0.1, xend = 1.0, yend = 0.1, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "both", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.885, y=0.125, label="SLRG", size = 3, color="black") +
  annotate(geom="text", x=0.8, y=0.950, label="SLCON = SWCON + SOIRG", size = 3, 
           color="black") +
  annotate(geom="text", x=0.8, y=0.9, label="SLRG = SWCON + SORG", size = 3, 
           color="black") +
  geom_segment(x = 0.575, y = 0.45, xend = 0.7, yend = 0.45, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "last", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.635, y=0.475, label = "KRGCL", size = 3, color="black") +
  geom_segment(x = 0.1, y = 1.0, xend = 0.25, yend = 1.0, lineend = "round", 
               linejoin = "round", size = 1, 
               arrow = arrow(end = "first", length = unit(0.075, "inches")), 
               colour = "black") + 
  annotate(geom="text", x=0.185, y = 0.975, label = "KROGCG", size = 3, color="black") +
  theme_bw()

p2
```
 
 

## `kr2p_ow() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SWCRIT`: critical water saturation, fraction.
+ `SOIRW`: irreducible oil saturation, fraction.
+ `SORW`: residual oil saturation, fraction.
+ `KRWIRO`: water relative permeability at irreducible oil.
+ `KROCW`: oil relative permeability at connate water.
+ `NW`: exponent term for calculating krw.
+ `NOW`: exponent term for calculating krow.
+ `NP`: number of saturation points in the table, the maximum acceptable value is 501.


## `kr2p_gl() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SOIRG`: irreducible oil saturation, fraction.
+ `SORG`: residual oil saturation, fraction.
+ `SGCON`: connate gas saturation, fraction.
+ `SGCRIT`: critical gas saturation, fraction.
+ `KRGCL`: gas relative permeability at connate liquid.
+ `KROGCG`: oil relative permeability at connate gas.
+ `NG`: exponent term for calculating krg.
+ `NOG`: exponent term for calculating krog.
+ `NP`: number of saturation points in the table, the maximum acceptable value is 501.


## `kr3p_StoneI_So() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SWCRIT`: critical water saturation, fraction.
+ `SOIRW`: irreducible oil saturation, fraction.
+ `SORW`: residual oil saturation, fraction.
+ `SOIRG`: irreducible oil saturation, fraction.
+ `SORG`: residual oil saturation, fraction.
+ `SGCON`: connate gas saturation, fraction.
+ `SGCRIT`: critical gas saturation, fraction.
+ `KRWIRO`: water relative permeability at irreducible oil.
+ `KROCW`: oil relative permeability at connate water.
+ `KRGCL`: gas relative permeability at connate liquid.
+ `KROGCG`: oil relative permeability at connate gas.
+ `NW`: exponent term for calculating krw.
+ `NOW`: exponent term for calculating krow.
+ `NG`: exponent term for calculating krg.
+ `NOG`: exponent term for calculating krog.
+ `NP`: number of saturation points in the two-phase relative permeability tables, the maximum acceptable value is 501. The number of data points in the three-phase relative permeability table is (0.5 * NP * (NP + 1)).



## `kr3p_StoneI_SwSg() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SWCRIT`: critical water saturation, fraction.
+ `SOIRW`: irreducible oil saturation, fraction.
+ `SORW`: residual oil saturation, fraction.
+ `SOIRG`: irreducible oil saturation, fraction.
+ `SORG`: residual oil saturation, fraction.
+ `SGCON`: connate gas saturation, fraction.
+ `SGCRIT`: critical gas saturation, fraction.
+ `KRWIRO`: water relative permeability at irreducible oil.
+ `KROCW`: oil relative permeability at connate water.
+ `KRGCL`: gas relative permeability at connate liquid.
+ `KROGCG`: oil relative permeability at connate gas.
+ `NW`: exponent term for calculating krw.
+ `NOW`: exponent term for calculating krow.
+ `NG`: exponent term for calculating krg.
+ `NOG`: exponent term for calculating krog.
+ `NP`: number of saturation points in the two-phase relative permeability tables, the maximum acceptable value is 501. The number of data points in the three-phase relative permeability table is (0.5 * NP * (NP + 1)).


## `kr3p_StoneII_So() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SWCRIT`: critical water saturation, fraction.
+ `SOIRW`: irreducible oil saturation, fraction.
+ `SORW`: residual oil saturation, fraction.
+ `SOIRG`: irreducible oil saturation, fraction.
+ `SORG`: residual oil saturation, fraction.
+ `SGCON`: connate gas saturation, fraction.
+ `SGCRIT`: critical gas saturation, fraction.
+ `KRWIRO`: water relative permeability at irreducible oil.
+ `KROCW`: oil relative permeability at connate water.
+ `KRGCL`: gas relative permeability at connate liquid.
+ `KROGCG`: oil relative permeability at connate gas.
+ `NW`: exponent term for calculating krw.
+ `NOW`: exponent term for calculating krow.
+ `NG`: exponent term for calculating krg.
+ `NOG`: exponent term for calculating krog.
+ `NP`: number of saturation points in the two-phase relative permeability tables, the maximum acceptable value is 501. The number of data points in the three-phase relative permeability table is (0.5 * NP * (NP + 1)).



## `kr3p_StoneII_SwSg() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SWCRIT`: critical water saturation, fraction.
+ `SOIRW`: irreducible oil saturation, fraction.
+ `SORW`: residual oil saturation, fraction.
+ `SOIRG`: irreducible oil saturation, fraction.
+ `SORG`: residual oil saturation, fraction.
+ `SGCON`: connate gas saturation, fraction.
+ `SGCRIT`: critical gas saturation, fraction.
+ `KRWIRO`: water relative permeability at irreducible oil.
+ `KROCW`: oil relative permeability at connate water.
+ `KRGCL`: gas relative permeability at connate liquid.
+ `KROGCG`: oil relative permeability at connate gas.
+ `NW`: exponent term for calculating krw.
+ `NOW`: exponent term for calculating krow.
+ `NG`: exponent term for calculating krg.
+ `NOG`: exponent term for calculating krog.
+ `NP`: number of saturation points in the two-phase relative permeability tables, the maximum acceptable value is 501. The number of data points in the three-phase relative permeability table is (0.5 * NP * (NP + 1)).



## `kr3p_Baker() arguments`

+ `SWCON`: connate water saturation, fraction.
+ `SWCRIT`: critical water saturation, fraction.
+ `SOIRW`: irreducible oil saturation, fraction.
+ `SORW`: residual oil saturation, fraction.
+ `SOIRG`: irreducible oil saturation, fraction.
+ `SORG`: residual oil saturation, fraction.
+ `SGCON`: connate gas saturation, fraction.
+ `SGCRIT`: critical gas saturation, fraction.
+ `KRWIRO`: water relative permeability at irreducible oil.
+ `KROCW`: oil relative permeability at connate water.
+ `KRGCL`: gas relative permeability at connate liquid.
+ `KROGCG`: oil relative permeability at connate gas.
+ `NW`: exponent term for calculating krw.
+ `NOW`: exponent term for calculating krow.
+ `NG`: exponent term for calculating krg.
+ `NOG`: exponent term for calculating krog.
+ `NP`: number of saturation points in the two-phase relative permeability tables, the maximum acceptable value is 501. The number of data points in the three-phase relative permeability table is (0.5 * NP * (NP + 1)).


## `Installation`

The `Rrelperm` can be installed from [CRAN](https://CRAN.R-project.org) with:


```{r, eval = FALSE, warning = FALSE}
install.packages("Rrelperm")
```


## `Two-phase Relative Permeability Examples`

#### Example 1:

```{r, fig.width = 6, fig.height= 4, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(magrittr)
library(tidyr)

rel_perm_ow <- kr2p_ow(SWCON = 0.13, SWCRIT = 0.13, SOIRW = 0.20, SORW = 0.2, 
                       KRWIRO = 0.45, KROCW = 1, NW = 4.25, NOW = 2.5, NP = 101)

rel_perm_ow_df <- as.data.frame(rel_perm_ow)
colnames(rel_perm_ow_df) <- c("Sw", "So", "Krw", "Krow")

head(rel_perm_ow_df, 10)

p1 <- rel_perm_ow_df %>% tidyr::pivot_longer(cols = c(Krw, Krow), values_to = "value", 
                                             names_to = "Kr") %>%
  ggplot(aes(x = Sw, y = value, col = Kr)) +
  geom_point() +
  labs(title = "water-oil relative permeability", x = "Sw", y = "Kr") +
  theme_bw()

p1
```


#### Example 2:

```{r, fig.width = 6, fig.height= 4, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(magrittr)

rel_perm_gl <- kr2p_gl(SWCON = 0.13, SOIRG = 0.12, SORG = 0.12, SGCON = 0.05, 
                       SGCRIT = 0.05, KRGCL = 0.35, KROGCG = 1, NG = 4.25, NOG = 2.5, 
                       NP = 101)

rel_perm_gl_df <- as.data.frame(rel_perm_gl)
colnames(rel_perm_gl_df) <- c("Sg", "Sl", "Krg", "Krog")

head(rel_perm_gl_df)

p2 <- rel_perm_gl_df %>% tidyr::pivot_longer(cols = c(Krg, Krog), values_to = "value", 
                                             names_to = "Kr") %>%
  ggplot(aes(x = Sg, y = value, col = Kr)) +
  geom_point() +
  labs(title = "gas-liquid relative permeability", x = "Sg", y = "Kr") +
  theme_bw()

p2
```


## `Three-phase Relative Permeability Examples`

#### Example 1: 

```{r, fig.width = 6, fig.height= 6, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(ggtern)
library(magrittr)

rel_perm_owg_st_i_swsg <- kr3p_StoneI_SwSg(SWCON = 0.13, SWCRIT = 0.13, SOIRW = 0.20, 
                                      SORW = 0.2, SOIRG = 0.1, SORG = 0.1, SGCON = 0, 
                                      SGCRIT = 0, KRWIRO = 0.45, KROCW = 1, KRGCL = 0.35, 
                                      NW = 4.25, NOW = 2.5, NG = 3, NOG = 2, 
                                      NP = 501)

rel_perm_owg_st_i_swsg <- as.data.frame(rel_perm_owg_st_i_swsg)
colnames(rel_perm_owg_st_i_swsg) <- c("Sw", "Sg", "So", "Kro")

head(rel_perm_owg_st_i_swsg, 10)

rel_perm_owg_st_i_swsg$Kro_range <- cut(rel_perm_owg_st_i_swsg$Kro, breaks = seq(0,1, by = 0.1),
                                   right = TRUE, include.lowest = FALSE)

rel_perm_owg_st_i_swsg_oil <- rel_perm_owg_st_i_swsg[rel_perm_owg_st_i_swsg$Kro > 0, ]

rel_perm_owg_st_i_swsg_oil <- droplevels(rel_perm_owg_st_i_swsg_oil)

p3 <- ggtern(data = rel_perm_owg_st_i_swsg_oil, aes(x = Sw, y = Sg, z = So, value = Kro)) +
  geom_point(aes(value = Kro, fill = Kro_range, color = Kro_range), size = 1, shape = 22) +
  scale_T_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_L_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_R_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  ggtitle("Three-phase Oil Relative Permeability - STONE I, SWSG") +
  theme_classic()

p3

```



#### Example 2: 

```{r, fig.width = 6, fig.height= 6, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(ggtern)
library(magrittr)

rel_perm_owg_st_i_so <- kr3p_StoneI_So(SWCON = 0.13, SWCRIT = 0.13, SOIRW = 0.20, 
                                      SORW = 0.2, SOIRG = 0.1, SORG = 0.1, SGCON = 0, 
                                      SGCRIT = 0, KRWIRO = 0.45, KROCW = 1, KRGCL = 0.35, 
                                      NW = 4.25, NOW = 2.5, NG = 3, NOG = 2, 
                                      NP = 501)

rel_perm_owg_st_i_so <- as.data.frame(rel_perm_owg_st_i_so)
colnames(rel_perm_owg_st_i_so) <- c("Sw", "Sg", "So", "Kro")

head(rel_perm_owg_st_i_so, 10)

rel_perm_owg_st_i_so$Kro_range <- cut(rel_perm_owg_st_i_so$Kro, breaks = seq(0,1, by = 0.1),
                                   right = TRUE, include.lowest = FALSE)

rel_perm_owg_st_i_so_oil <- rel_perm_owg_st_i_so[rel_perm_owg_st_i_so$Kro > 0, ]

rel_perm_owg_st_i_so_oil <- droplevels(rel_perm_owg_st_i_so_oil)

p4 <- ggtern(data = rel_perm_owg_st_i_so_oil, aes(x = Sw, y = Sg, z = So, value = Kro)) +
  geom_point(aes(value = Kro, fill = Kro_range, color = Kro_range), size = 1, shape = 22) +
  scale_T_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_L_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_R_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  ggtitle("Three-phase Oil Relative Permeability - STONE I, SO") +
  theme_classic()

p4

```



#### Example 3: 

```{r, fig.width = 6, fig.height= 6, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(ggtern)
library(magrittr)

rel_perm_owg_st_ii_swsg <- kr3p_StoneII_SwSg(SWCON = 0.13, SWCRIT = 0.13, SOIRW = 0.20, 
                                      SORW = 0.2, SOIRG = 0.1, SORG = 0.1, SGCON = 0, 
                                      SGCRIT = 0, KRWIRO = 0.45, KROCW = 1, KRGCL = 0.35, 
                                      NW = 4.25, NOW = 2.5, NG = 3, NOG = 2, 
                                      NP = 501)

rel_perm_owg_st_ii_swsg <- as.data.frame(rel_perm_owg_st_ii_swsg)
colnames(rel_perm_owg_st_ii_swsg) <- c("Sw", "Sg", "So", "Kro")

head(rel_perm_owg_st_ii_swsg, 10)

rel_perm_owg_st_ii_swsg$Kro_range <- cut(rel_perm_owg_st_ii_swsg$Kro, breaks = seq(0,1, by = 0.1),
                                   right = TRUE, include.lowest = FALSE)

rel_perm_owg_st_ii_swsg_oil <- rel_perm_owg_st_ii_swsg[rel_perm_owg_st_ii_swsg$Kro > 0, ]

rel_perm_owg_st_ii_swsg_oil <- droplevels(rel_perm_owg_st_ii_swsg_oil)

p5 <- ggtern(data = rel_perm_owg_st_ii_swsg_oil, aes(x = Sw, y = Sg, z = So, value = Kro)) +
  geom_point(aes(value = Kro, fill = Kro_range, color = Kro_range), size = 1, shape = 22) +
  scale_T_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_L_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_R_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  ggtitle("Three-phase Oil Relative Permeability - STONE II, SWSG") +
  theme_classic()

p5

```


#### Example 4: 

```{r, fig.width = 6, fig.height= 6, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(ggtern)
library(magrittr)

rel_perm_owg_st_ii_so <- kr3p_StoneII_So(SWCON = 0.13, SWCRIT = 0.13, SOIRW = 0.20, 
                                      SORW = 0.2, SOIRG = 0.1, SORG = 0.1, SGCON = 0, 
                                      SGCRIT = 0, KRWIRO = 0.45, KROCW = 1, KRGCL = 0.35, 
                                      NW = 4.25, NOW = 2.5, NG = 3, NOG = 2, 
                                      NP = 501)

rel_perm_owg_st_ii_so <- as.data.frame(rel_perm_owg_st_ii_so)
colnames(rel_perm_owg_st_ii_so) <- c("Sw", "Sg", "So", "Kro")

head(rel_perm_owg_st_ii_so, 10)

rel_perm_owg_st_ii_so$Kro_range <- cut(rel_perm_owg_st_ii_so$Kro, breaks = seq(0,1, by = 0.1),
                                   right = TRUE, include.lowest = FALSE)

rel_perm_owg_st_ii_so_oil <- rel_perm_owg_st_ii_so[rel_perm_owg_st_ii_so$Kro > 0, ]

rel_perm_owg_st_ii_so_oil <- droplevels(rel_perm_owg_st_ii_so_oil)

p6 <- ggtern(data = rel_perm_owg_st_ii_so_oil, aes(x = Sw, y = Sg, z = So, value = Kro)) +
  geom_point(aes(value = Kro, fill = Kro_range, color = Kro_range), size = 1, shape = 22) +
  scale_T_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_L_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_R_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  ggtitle("Three-phase Oil Relative Permeability - STONE II, SO") +
  theme_classic()

p6

```


#### Example 5: 

```{r, fig.width = 6, fig.height= 6, fig.align = "center", warning = TRUE}
library(Rrelperm)
library(ggplot2)
library(ggtern)
library(magrittr)

rel_perm_owg_bk <- kr3p_Baker(SWCON = 0.13, SWCRIT = 0.13, SOIRW = 0.20, 
                                      SORW = 0.2, SOIRG = 0.1, SORG = 0.1, SGCON = 0, 
                                      SGCRIT = 0, KRWIRO = 0.45, KROCW = 1, KRGCL = 0.35, 
                                      NW = 4.25, NOW = 2.5, NG = 3, NOG = 2, 
                                      NP = 501)

rel_perm_owg_bk <- as.data.frame(rel_perm_owg_bk)
colnames(rel_perm_owg_bk) <- c("Sw", "Sg", "So", "Kro")

head(rel_perm_owg_bk, 10)

rel_perm_owg_bk$Kro_range <- cut(rel_perm_owg_bk$Kro, breaks = seq(0,1, by = 0.1),
                                   right = TRUE, include.lowest = FALSE)

rel_perm_owg_bk_oil <- rel_perm_owg_bk[rel_perm_owg_bk$Kro > 0, ]

rel_perm_owg_bk_oil <- droplevels(rel_perm_owg_bk_oil)

p7 <- ggtern(data = rel_perm_owg_bk_oil, aes(x = Sw, y = Sg, z = So)) +
  geom_point(aes(value = Kro, fill = Kro_range, color = Kro_range), size = 1, shape = 22) +
  scale_T_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_L_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  scale_R_continuous(limits = c(0.0,1.0), breaks = seq(0,1,by = 0.05)) +
  ggtitle("Three-phase Oil Relative Permeability - Baker's Linear") +
  theme_classic()

p7

```


## `References`
