---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **flexpolyline** R package provides a binding to the
[C++ implementation](https://github.com/heremaps/flexible-polyline/tree/master/cpp) of the
flexible polyline encoding by [HERE](https://github.com/heremaps/flexible-polyline).
The flexible polyline encoding is a lossy compressed representation of a list of
coordinate pairs or coordinate triples.

The encoding is achieved by:

(1) Reducing the decimal digits of each value;
(2) encoding only the offset from the previous point;
(3) using variable length for each coordinate delta; and
(4) using 64 URL-safe characters to display the result.
The felxible polyline encoding is a variant of the [Encoded Polyline Algorithm Format](https://developers.google.com/maps/documentation/utilities/polylinealgorithm) by Google.

**Note:**

* Decoding gives reliable results up to a precision of 7 digits.
The tests are also limited to this range.
* The order of the coordinates (lng, lat) does not correspond to the original C++ implementation (lat, lng).
This enables simple conversion to `sf` objects, without reordering the columns.
* The encoding is lossy, this means the encoding process could reduce the precision of your data.

## Installation
You can install the released version of **flexpolyline** from [CRAN](https://CRAN.R-project.org/package=flexpolyline) with:

```{r, eval=FALSE}
install.packages("flexpolyline")
```

Install the development version from [GitHub](https://github.com/munterfinger/flexpolyline) with:

```{r, eval=FALSE}
remotes::install_github("munterfinger/flexpolyline")
```

## C++ binding
Encoding and decoding in R is straight forward by using `encode()` and `decode()`.
These functions are binding to the flexpolyline C++ implementation and have similar arguments
and return values as their counterparts (`hf::encode_polyline` and `hf::decode_polyline`):

```{r cpp_encode_decode}
library(flexpolyline)

(line <- matrix(
  c(8.69821, 50.10228, 10,
    8.69567, 50.10201, 20,
    8.69150, 50.10063, 30,
    8.68752, 50.09878, 40),
  ncol = 3, byrow = TRUE
))

(encoded <- encode(line))

(decoded <- decode(encoded))
```

## Simple feature support
A common way to deal with spatial data in R is the
[sf](https://CRAN.R-project.org/package=sf) package, which is
built on the concept of simple features. The functions `encode_sf()` and
`decode_sf()` provide an interface that support the encoding of sf objects:

```{r sf_encode_decode}
(sfg <- sf::st_linestring(line, dim = "XYZ"))

(encoded <- encode_sf(sfg))

(decoded <- decode_sf(encoded))
```

## References

* [Flexible Polyline Encoding by HERE](https://github.com/heremaps/flexible-polyline)
* [Encoded Polyline Algorithm Format](https://developers.google.com/maps/documentation/utilities/polylinealgorithm)
* [Simple Features for R](https://CRAN.R-project.org/package=sf)
* Inspired by the [googlePolylines](https://github.com/SymbolixAU/googlePolylines) package
