---
title: "GenEst - A Tutorial with Wind Examples"
author: "Daniel Dalthorp"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GenEst - 1. A Tutorial with Wind Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 120)
```

```{r, include=FALSE}
library(GenEst)
vers <- packageVersion("GenEst")
today <- Sys.Date()
```

## Introduction: Tutorial with Examples
This tutorial provides an introduction to the array of command line tools
**GenEst** (`r vers`) provides for estimating bird and/or bat mortality and
detection probabilities and wind power facilities. The approach is to walk
through analyses of realistic but simulated data sets representing studies
of bird and bat mortality at a wind power facilities.

The general steps in the analysis are:

1. Construct a model for Searcher Efficiency
2. Construct a model for Carcass Persistance
3. Estimate mortality
4. Specify the type of summary desired (e.g., mortality by species and season)

Data required for a full analysis include results of searcher efficiency trials,
results of carcass persistence trials, search schedules for all units searched,
the search coverage within each unit, and results of periodic carcass surveys.
More information about the kinds of data required can be found in the User Guide,
which can be found at `https://code.usgs.gov/ecosystems/GenEst/-/releases` in
the "For more info" section. For convenience, data required in this tutorial are
available in R as packaged lists, which can readily be loaded as described
below. Alternatively, the data may be downloaded into zipped .csv files from the
`code.usgs.gov` page referenced above.

To perform the analyses illustrated in the tutorial, begin by starting R and
loading GenEst from the command line: `library(GenEst)`

## Example 1: Estimating Bat Mortality from Searches on Roads and Pads
Searcher efficiency and carcass persistence would be expected to vary with
carcass size (sparrow, eagle, bat), ground characteristics (road & pad,
cleared field, groud texture), season, etc. In this first example, we limit
the analysis to one carcass size (`bat`) and one ground visibility class
(`RP` = road and pad). A more complicated scenario is analyzed in example 2.

The required data is stored in `wind_RPbat`, a list of data frames with results
for searcher efficiency (`SE`) and carcass persistence trials (`CP`), search
schedules for all turbines (`SS`), the search coverage or density weighted
proportion (`DWP`) of area searched at each turbine (i.e., the fraction of
carcasses expected to fall in the search plots), and carcass observation (`CO`)
data.

Load the full data set into R:
```{r}
data(wind_RPbat)
names(wind_RPbat)
```
To streamline the notation, extract the data from the `wind_RPbat` list into its
components:
```{r}
data_SE <- wind_RPbat$SE
data_CP <- wind_RPbat$CP
data_SS <- wind_RPbat$SS
data_DWP <- wind_RPbat$DWP
data_CO <- wind_RPbat$CO
```

```{r, include = FALSE}
daterange <- range(data_SS$SearchDate)
seasons <- paste(unique(data_SS$Season), collapse = ', ')
```
### Searcher Efficiency (`SE`)
Searcher efficiency trials were conducted on roads and pads, with a total of
`r  dim(data_SE)[1]` fresh carcasses placed in the field over the course of
the entire monitoring period, evenly divided among seasons (`r seasons`). 
Carcasses that were later discovered by search teams during the course of 
normal carcass surveys were removed from the field. Carcasses were left in 
the field for up to 5 searches after carcass placement.

Results of the SE field trials are stored in the `data_SE` data frame:
```{r}
head(data_SE)
```
Columns `s1, s2, ..., s5` show the fate of carcass `pkID` on the 1st, 2nd, ...
5th searches after the carcass was placed. A 1 indicates that the carcass was
discovered, a 0 indicates that the carcass was present but not discovered, and
NA indicates that the carcass was no longer present for discovery or no search
was conducted.

### Carcass Persistence (`CP`)
Carcass persistence trials were conducted on roads and pads, with a total of
`r  dim(data_CP)[1]` fresh carcasses placed in the field over the course of
the entire monitoring period, evenly divided among seasons (`r seasons`). 
Carcasses were checked approximately 1, 2, 3, 4, 7, 10, 14, 21, and 28
days after placement in the field. Exact times were entered as decimal
fractions of days after placement.

Results of the SE field trials are stored in the `data_CP` data frame:
```{r}
head(data_CP)
```

Exact persistence times are not known, but a carcass that was present at one
check and absent at the next check is assumed to have been removed at some
point in the interval. The left endpoint of the interval was entered as
`LastPresent` and the right endpoint as `FirstAbsent`. For carcasses that had
not been scavenged by the end of the study, `LastPresent` is the time of the
last check and `FirstAbsent` is `Inf`. For carcasses whose removal time is known
exactly (e.g., scavenging was recorded by camera), `LastPresent = FirstAbsent`.
The `Season` column gives the season at the time the carcass was placed in the
field.

### Search Schedules (`SS`)
Carcass searches were conducted on roads and pads within a 120 m radius from 
all `r dim(data_DWP)[1]` turbines at our fictitious site. Monitoring began on
`r daterange[1]` and continued through `r daterange[2]`. Searches spanned
`r length(unique(data_SS$Season))` seasons: `r seasons`. Search intervals varied
by turbine and by time of year, ranging from daily searches at some turbines in
the fall and searches once every 12 days in the spring at some other turbines.
Search schedules for all turbines are stored in `data_SS`, which is a data frame
with a column for search dates (including all dates that any turbine was
searched); a column of 0s and 1s for each turbine, indication whether it was
searched on the given date; and zero or more optional columns giving additional
information about the date (e.g., season).

```{r}
head(data_SS[, 1:10])
```

Note that we have only displayed a few of the turbine columns - there are 100 
turbine columns altogether (t1, ..., t100).


### Density Weighted Proportion (`DWP`)
The density-weighted proportion (`DWP`) is the expected fraction of carcasses
that fell in the searched area. Carcass density is not the same at all distances
from a turbine, but typically rises over a short distance then decreases
eventually to 0. Searches were conducted on roads and pads within a 120 m
radius from all 100 turbines to provide sufficient data with which to model the
change in density with distance and from this, accurately calculate the fraction
of all carcasses that are expected to land on road or pad surrounding each
turbine (density-weighted proportion or DWP). The configuration of the
roads and pads differs among turbines, hence the DWP must be calculated for each
turbine. DWPs for bats at each turbine are stored in `data_DWP`, which is a data
frame with a column for turbine name (note that turbine names also must be
included among column names in `data_SS`, which gives the search schedule at
each turbine) and a column of DWP labeled `bat`. In other studies where, for example,
mortality of birds might be of interest, `DWP` would be expected vary with
carcass size or species (e.g., the spatial distributions of bats and large birds
around a turbine would likely differ from one another). In that case, each
carcass size class (`large`, `medium`, `small`, `bat`) would have its own column.

```{r}
head(data_DWP)
```

### Carcass Observations (`CO`)
Information about each (non-trial) carcass observed during searches is stored in
`CO_data` which is a data frame with at least 4 columns: carcass ID, the turbine
(or unit) at which it was found, the date it was found and its distance from the
turbine center. In `data_CO` we also have turbine type, species, and species
group variables by which we will later summarize mortality estimates.
```{r}
head(data_CO)
```

### Estimating Searcher Efficiency and Carcass Persistence Parameters
Searcher efficiency and carcass persistence parameters are estimated by fitting
models using functions `pkm` and `cpm` (carcass persistence model) which are
patterned after familar R functions such as `lm` and `glm`. The "pk" in `pkm`
refers to GenEst's model of searcher efficiency which includes two parameters:
`p`, which is the initial searcher efficiency for carcasses on the first search
after they have arrived, and `k`, which is a parameter governing the decrease
in searcher efficiency in later searches. In this relatively simple example,
our SE and CP field trials were conducted for one carcass size (bat) on one
type of terrain (roads and pads) in three seasons (spring, summer, fall). The
only potential predictor variable we have is `Season`, which is entered as a
column in both `data_SE` and `data_CP`.

Searcher efficiency is the probability of detection of a carcass that is present
in the searched area at the time of search. Searcher efficiency typically
decreases with carcass age because older carcasses tend to become harder to find
as they accumulate dust or debris, fall deeper into vegetation, get blown
against objects or into holes, decay, or get partially scavenged. In addition,
carcasses missed in one search tend to be more likely to be missed in subsequent
searches because the relatively easy-to-find carcasses are preferentially
removed in the first searches after carcass arrival, leaving mostly the
harder-to-find carcasses available in subsequences searches. GenEst accounts
for a non-constant searcher efficiency using two parameters, `p` (searcher
efficiency on the first search after carcass arrivals) and `k` (proportional
change in searcher efficiency with each successive search). The `k` parameter
can be estimated from field trials if carcasses that are not discovered in the
first search after arrival are left in the field for possible discovery in
later searches.

`p` and `k` may both depend on covariates such as season, visibility class, or
carcass size, and GenEst allows for them to be modeled as functions of different
covariate combinations.

```{r}
model_SE <- pkm(p ~ Season, k ~ 1, data = data_SE)
class(model_SE)
model_SE
```

NOTE: The `pkm` family of functions by default interprets columns with names
that begin with and "s" or "S" and end with a number contain search results data
(carcass found = 0, not found = 1). A user can override the auto-parsing by
explicitly listing the names of the search data columns in a vector of character
strings in the `obsCol` argument.

The probability of a carcass persisting a given length of time without being
removed by scavengers (or other factors) is modeled as a Weibull, lognormal,
loglogistic, or exponential distribution. Like the `p` and `k` parameters for
searcher efficiency, the location and scale parameters (Therneau 2015) of the
persistence distribution may depend on covariates. GenEst allows for them to be
modeled as separate functions of predictor combinations.

```{r}
model_CP <- cpm(l ~ Season, s ~ Season, data = data_CP, dist = "weibull",
  left = "LastPresent", right = "FirstAbsent")
class(model_CP)
model_CP
```
The model summary shows descriptive statistics for the cellwise estimates of
the `l` and `s` parameters. The `location` and `scale` parameterization is
common in survival analysis, but the `pda` and `pdb` parameterization (Dalthorp
and Huso, 2014) is also shown. These parameterizations are convenient to work
with in statistical calculations but are not as convenient for giving users
quick insight into the distributions, so a third set of summary statistics
about the fitted distributions is also given. Namely, the median persistence
time and the `r` statistic, which is the probability that a carcass persists
until the first search after arrival (assuming uniformly distributed arrival
times within the interval). Clearly, `r` depends on the length of the search
interval, and the table shows `r` for intervals of 1, 3, 7, 14, and 28 days. A
rough, back-of-the-envelope calculation for the probability of observing a
carcass that arrives at a site during the monitored period would be
`DWP * r * p * f`, which is `DWP` = fraction of carcasses that arrive in the
area searched at a unit, `r` = fraction of carcasses that persist until a
search, `p` = fraction of carcasses found on the first search after arrival
(given that they persisted), and `f` = fraction of carcasses that arrive at
the units searched.

In other scenarios we might consider other predictors, like the visibility of
the ground searched or the search team. We might also be interested in carcasses
of different sizes (e.g., large, medium, and small birds instead of or in
addition to bats). We are not restricted to using the same predictors of both SE
and CP. The modeling complexity increases with each additional predictor, but,
in theory, any number of predictors can be used. The only rule is that
sufficient numbers of trial carcasses must be placed in each cell combination
of factor levels among the selected predictors. For example, if we were to place
15 carcasses for each cell for predictors that include season (spring, summer,
fall, winter), size (S, M, L, B), visibility (RP, M, D), search team (dogs,
humans), and turbine type (small, medium, large), we'd need
15 x 4 x 4 x 3 x 2 x 2 = 2880 carcasses. Typically, the number of predictors is
limited to a few key variables.

### Mortality Estimation
Each carcass's contribution to the total mortality in each search interval is
estimated using the `estM` function.

```{r, fig.width = 7, fig.height = 5, fig.align = 'center'}
Mhat <- estM(nsim = 1000, data_CO = data_CO, data_SS = data_SS,
  data_DWP = data_DWP, model_SE = model_SE, model_CP = model_CP,
  unitCol = "Turbine", COdate = "DateFound")

summary(Mhat)
plot(Mhat)
```

Mortality estimates may be partitioned or split into desired categories, such as
species, season, or turbine type. Splits may be performed according to
characteristics of the carcasses or where they were found (e.g., species,
turbine or other variable found in `data_CO`) or when they were found (e.g.,
season or other variable associated with search schedule and found in `data_SS`,
or a vector of specific times).

Mortality by `Species` (a CO split because it is a column in the CO file):
```{r, fig.width = 5, fig.height = 5, fig.align = 'center'}
M_species <- calcSplits(M = Mhat, split_CO = "Species", data_CO = data_CO)
summary(M_species)
plot(M_species)
```

Mortality estimates may also be split by temporal variables that are represented
as columns in `data_SS` or as numeric vectors spanning the monitoring
season (from day 0 to length of monitoring season). If several temporal splits
are to be calculated, creating a specially formatted `prepSS` object for the
search schedule can streamline the calculations.
```{r}
SSdat <- prepSS(data_SS)
```

Mortality by `Season` (an SS split because it is a column in the SS file):
```{r Season Split, fig.width = 6.5, fig.height = 5.25, fig.align = 'center'}
M_season <- calcSplits(M = Mhat, split_SS = "Season", data_SS = SSdat,
  split_CO = NULL,  data_CO = data_CO)
summary(M_season)
plot(M_season)
```

Mortality by month (a temporal split that spans the monitoring period):
```{r Temporal Split, fig.width = 7, fig.height = 5, fig.align = 'center'}
M_month <- calcSplits(M = Mhat, split_time = seq(0, max(SSdat$days), by = 28),
  data_SS = SSdat, data_CO = data_CO)
summary(M_month)
plot(M_month)
```
Temporal splits that divide the monitoring season into separate time intervals
(like season or month) can be plotted as the number per interval
(`rate = FALSE`, which is the default arg in `calcSplits`) or the number per
unit time (`rate = TRUE`).
```{r Time unit Split, fig.width = 7, fig.height = 5, fig.align = 'center'}
M_various_times <- calcSplits(M = Mhat,
  split_time = c(seq(0, 90, by = 15), 120, 150, seq(155, 200, by = 5)),
  data_SS = SSdat, data_CO = data_CO)
plot(M_various_times)
plot(M_various_times, rate = TRUE)
```

Finally, splits can be calculated for combinations of splitting covariates,
like species by season or species group by turbine type. No more than two
splitting covariates may be used in one call to `calcSplits` and at most one
temporal split may be used (whether it is an SS split or a vector of times).

```{r Species and Season, fig.width = 4, fig.height = 6, fig.align = 'center'}
M_species_by_season <- calcSplits(M = Mhat,
  split_CO = "Species", data_CO = data_CO,
  split_SS = "Season", data_SS = SSdat)
plot(M_species_by_season)
```

## Example 2: Estimating Bird and Bat Mortality from Searches on Varied Ground
Thorough searches out to a radius of 60 m from each turbine were conducted at 23
out of 100 turbines. The searched area was divided into three visibility classes
(`RP`, `M`, `D`) according the difficulty of finding carcasses.

Searcher efficiency and carcass persistence would be expected to vary with
carcass size (sparrow, eagle, bat), ground characteristics (road & pad,
cleared field, vegetation type), season, etc. In this example, we perform a full
analysis of scenario with four classes of carcass (`lrg`, `med`, `sml`, and
`bat`), three visibility classes (difficult = `D`,  moderate = `M`, and road &
pad = `RP`), and three seasons (`spring`, `summer`, and `fall`).

The required data is stored in `wind_cleared`, a list of data frames with results
for searcher efficiency (`SE`) and carcass persistence trials (`CP`), search
schedules for all turbines (`SS`), the search coverage or density weighted
proportion (`DWP`) of area searched at each turbine (i.e., the fraction of
carcasses expected to fall in the search plots), and carcass observation (`CO`)
data.

Load the full data set into R:
```{r}
data(wind_cleared)
names(wind_cleared)
```
To streamline the notation, extract the data from the `wind_cleared` list into
its components:
```{r}
data_SE <- wind_cleared$SE
data_CP <- wind_cleared$CP
data_SS <- wind_cleared$SS
data_DWP <- wind_cleared$DWP
data_CO <- wind_cleared$CO
```
### Searcher Efficiency and Carcass Persistence Trials
In searcher efficiency and carcass persistence trials, 15 trial
carcasses were placed in each combination of visibility class (`D`, `M`, `RP`),
season (`spring`, `summer`, `fall`), and size class (`lrg`, `med`, `sml`, `bat`).
Data formats are like those of example 1:
```{r}
head(data_SE)
head(data_CP)
```

### Searcher Efficiency Modeling
With 36 combinations of covariate levels (3 visibilities x 3 seasons x 4 sizes)
and two parameters (`p` and `k`), the number of possible models to consider for
searcher efficiency is unwieldy using simple calls to `pkm`, but `pkm` has
powerful model building and model selection capabilities that can be accessed
via the arg list: `allCombos` and `sizeCol`, which are discussed below.

When `allCombos = TRUE`, `pkm` fits the set of submodels of the given covariate
combinations, including the full model, the null model, and
everything in between. For example, if the parameter models are
`p ~ Visibility * Season` and `k ~ Visibility`, `pkm` with `allCombos = TRUE`
would fit all combinations of possible `p` models (`p ~ Visibility * Season`,
`p ~ Visibility + Season`, `p ~ Visibility`, `p ~ Season`, and `p ~ 1`) and
possible `k` models (`k ~ Visibility` and `k ~ 1`), or 10 models in all.

Carcasses in different size classes would be expected to have different searcher
efficiency and carcass persistence parameters and would even be likely to be
affected by covariates in different ways. When analyzing data with carcasses in
different size classes, it is recommended that separate CP and SE models be fit
for size classes separately. This can be accomplished using the `sizeCol`
argument in `pkm` and `cpm`, where `sizeCol` gives the name of the column that
gives the carcass size classes in `data_SE`.

```{r}
pkModels <- pkm(p ~ Visibility * Season, k ~ Visibility * Season, data = data_SE,
  allCombos = TRUE, sizeCol = "Size")
class(pkModels)
names(pkModels)
```
When `allCombos = TRUE` and `sizeCol` is defined, `pkm` returns a list of sets
of models for each size class. The sets of models for each size class include
the full spectrum of models that can be constructed using simple combinations
of the covariates.
```{r}
names(pkModels[["sml"]])
class(pkModels[["sml"]])
```

To estimate mortality, one model for each size class must be selected from the
long list of models fit. GenEst provides several tools for guiding the selection.
First, the models can be listed by AICc, which gives a score for the quality of
the model for the given data. Complicated models that use many parameters may
fit the data more closely than a simpler model but are penalized because of
their complexity and relative instability. The scores have meaning only in
comparison with other models'. AICc provides a rough but useful guide for model
selection, but should in no way be relied upon as definitive. Its utility is in
identifying relatively poor models and in narrowing the choice of plausible
models to a manageable number.

The `aicc` function lists the fitted models in order of ${\small \Delta}$AICc
for each size class (if applicable). For this discussion, we will focus on
`sml` only, but for a full analysis, all size classes would need to be similarly
analyzed.

```{r}
aicc(pkModels[["sml"]])
```
Preference should normally be given to models with ${\small \Delta}$AICc less
than 6 or 7. Models with differences of less than 3 or 4 are generally
considered indistinguishable by this measure. Choices among such models should
be based on other criteria.

Diagnostic plots can be used to identify potential problems with model fits and
to help distinguish between models with similar AICc scores. The `plot` function
is defined for `pkm` (one model) and `pkmSet` (set of models for a given size
class) objects. For example, `plot(pkModels[["sml"]][[1]])` would produce the
single figure for the first model for the `sml` size class.
`plot(pkModels[["sml"]])` would create plots for each model fit for the `sml`
size class. To plot a specific single model from the full set,
use the `specificModel` argument. For example, diagnostic plots for the model
with the lowest AICc score are shown below:

```{r, fig.show = "hold", fig.width = 7, fig.height = 7}
plot(pkModels[["sml"]], specificModel = "p ~ Visibility; k ~ 1")
```

The top row shows box plots of estimated `p` and `k` parameters for all cells
(i.e., combinations of covariate levels, like `D.fall` for difficult visibility
in the fall) for both the selected model (blue) and the full model (gray). With
the full model, the fits for each cell are based solely on data from that
specific cell. The advantage is that each cell's estimates are untainted by data
from other cells. The disadvantage is that the sample size for each estimate
is relatively small and the error bars large. In the reduced models, estimates
for one cell borrow strength from estimates in related cells. This gives smaller
error bars but can lead to errors if the model structure does not properly
reflect the dependence of searcher efficiency parameters on cell
characteristics.

In the figure, the estimates of `p` from the selected model are markedly less
variable than the estimates from the full model, while the locations of the
boxes are very similar for the two models. Thus, this selected model (the one
with the lowest AICc) appears to be an improvement over the full cell model for
estimating `p`.

The ${\small \Delta}$AICc for the full model is 12.52, which indicates a serious
deficiency in comparison to the model with the best fit,
`"p ~ Visibility; k ~ 1"`. The boxplots for `k` highlight one particular problem
with the fit of the reference model. In some of the cells, the boxes extend from
0 to 1, which suggests that the reference model is unable to estimate `k` for
the given cell. Selecting a simpler model for `k` often remedies this problem.
If all models display this 0-1 phenomenon, a fixed `k` of 1 is appropriate if
a smaller proportion of carcasses was found on the first search occasion than on
later searches.

By comparison, the model with the highest ${\small \Delta}$AICc (=
`{r max(aicc(pkModels[["sml"]]))}` routinely estimates `p` and `k` either well
above or well below the reference model but has fairly tight error bars--bad
estimates but quite confident about them!

```{r, fig.show = "hold", fig.width = 7, fig.height = 7}
plot(pkModels[["sml"]], specificModel = "p ~ Season; k ~ Visibility * Season")
```

A similar model selection exercise gives the same form of model
(`p ~ Visibility; k ~ 1`) for large birds, medium birds, and bats. These can all
be collated into a list for later analysis of detection probabilities and
mortality rates.

```{r}
pkMods <- list(
  sml = pkModels[["sml"]][["p ~ Visibility; k ~ 1"]],
  med = pkModels[["med"]][["p ~ Visibility; k ~ 1"]],
  lrg = pkModels[["lrg"]][["p ~ Visibility; k ~ 1"]],
  bat = pkModels[["bat"]][["p ~ Visibility; k ~ 1"]]
)
```

### Carcass Persistence Modeling
The work flow for carcass persistence modeling is similar to that for searcher
efficiency except that in addition to selecting covariates for two different
parameters (location = `l` and scale = `s`), there are four model forms to
choose from: Weibull, lognormal, loglogistic, and exponential. 

```{r}
cpModels <- cpm(
  l ~ Visibility * Season, s ~ Visibility * Season,
  data = data_CP, left = "LastPresent", right = "FirstAbsent",
  dist = c( "weibull", "lognormal", "loglogistic", "exponential"),
  allCombos = TRUE, sizeCol = "Size"
)
```
The list of models is long:
```{r}
lapply(aicc(cpModels), nrow)
aicc(cpModels[["sml"]])
```
It is not uncommon to see the fits for the exponential distribution at the
bottom of the AIC list. The exponential distribution has only one parameter and
does not have nearly as much flexibility as the others, which each have
two-parameters. An implicit assumption of the exponential model is that the
scavenging rate is constant, regardless of carcass age. When that assumption is
not met, the exponential provides an inferior fit. 

To compare among a set of cp models with plausible AICs, use, for example,
`plot(cpModels[["sml"]])` to browse through all the models or
`plot(cpModels[["sml"]][["dist: lognormal; p ~ Visibility; k ~ 1"]])` to view
a single model specified by name. It can be seen from the AICc table
(`aicc(cpModels[["sml"]])`) that the top 10 models according to AICc are:

```{r}
cp_smlCandidates <-
    names(cpModels[["sml"]])[c(25, 24, 15, 19, 20, 10, 14, 22, 74, 21)]
cp_smlCandidates
```

These can be compared in graphs as follows:

```{r, eval = F}
plot(cpModels[["sml"]], specificModel = cp_smlCandidates)
```

The figure shows the raw persistence data (fraction of carcasses remaining after
the given time) for each cell, as a black stair case with Kaplan-Meier
confidence intervals as dashed lines. In addition, the fitted curves for each of
the distributions are shown in color, with the `specificModel` distribution
having a thicker than the others. The two-parameter models (Weibull, lognormal,
and loglogistic) tend to be very similar and a relatively close fit to the data.
The exponential model tends to be somewhat removed from the others. Clicking on
the graphing window brings up the next set of figures.

We're looking for a good fit between the selected model and the data in as many
cells as possible. The first several models among the `cp_smlCandidates` seems
to provide a reasonably good fit in all cells, although there seems to be a
trade-off between fitting the RP.spring cell well or fitting the RP.summer cell
well. `Visibility` occurs frequently in the top models, while `Season` appears
more frequently in the bottom models. This indicates that `Season` is probably
not a strong predictor of carcass persistence, while `Visibility` is.

Selecting the top AICc model for use in mortality estimation:
```{r}
cp_sml <- cpModels[["sml"]][[cp_smlCandidates[1]]]
```

Following a similar model selection process for the other size classes, we
select the following:

```{r}
cp_med <- cpModels[["med"]][["dist: weibull; l ~ Visibility; s ~ Season"]]
cp_lrg <- cpModels[["lrg"]][["dist: exponential; l ~ Visibility + Season; NULL"]]
cp_bat <- cpModels[["bat"]][["dist: weibull; l ~ Visibility + Season; s ~ 1"]]
```

NOTE: For large carcasses, the exponential distribution was at the top of the AICc
list. The exponential requires only one parameter (`l`), so no scale parameter
is provided in the model.

Again, we collate the models into a list for later analysis of detection
probabilities and mortality rates.

```{r}
cpMods <- list(
  sml = cp_sml,
  med = cp_med,
  lrg = cp_lrg,
  bat = cp_bat
)
```

### Mortality Estimation
Each carcass's contribution to the total mortality in each search interval is
estimated using the `estM` function. The function call is largely similar that
used in the simple scenario discussed in example 1. However, there are some
important differences. First, `model_SE` and `model_CP` are lists of models, one
element for each size class. In addition, the name of the size class variable is
provided as `sizeCol = "Size"`. Finally, the `frac` argument represents the
sampling fraction or the fraction of carcasses expected to fall at the units
that were searched. In this example, 23 out of 100 turbines were searched, and
`frac` is set equal to 0.23 under the assumption that the mortality rates at the
unsearched turbines did not differ substantially from the rates at the searched
turbines.

```{r Mhat plot, fig.height = 4, fig.width = 7, fig.align = 'center'}
Mhat <- estM(nsim = 1000, data_CO = data_CO, data_SS = data_SS, frac = 0.23,
  data_DWP = data_DWP, model_SE = pkMods, model_CP = cpMods,
  sizeCol = "Size", unitCol = "Turbine", COdate = "DateFound")

summary(Mhat)
plot(Mhat)
```

This estimate is for the total number of fatalities among all size classes
combined, from hummingbirds and bats to eagles and may be too vague to be very
useful. Fortunately, mortality estimates
may be partitioned or split into desired categories, such as species, size, or
season. Splits may be performed according to characteristics
of the carcasses or where they were found (e.g., species, turbine or other
variable found in `data_CO`) or when they were found (e.g., season or other
variable associated with search schedule and found in ``data_SS``, or a vector of
specific times).

Carcasses were categorized not only by size but also by species, species group,
the type of turbine they were found at, the visibility class of the ground
where they were found, and distance from nearest turbine.

Although species groups may extend across different size classes (e.g.,
`raptors` could include kestrels, red-tailed hawks, and golden eagles;
`passerines` could include sparrows and ravens), splits according to species
group can easily be accomplished:

```{r Species Group Plot, fig.height = 5, fig.width = 5, fig.align = 'center'}
M_speciesGroup <- calcSplits(M = Mhat,
  split_CO = "SpeciesGroup",  data_CO = data_CO)
summary(M_speciesGroup)
plot(M_speciesGroup)
```

Split by species and season:
```{r Split Species and Season, fig.height = 12, fig.width = 4, fig.align = 'center'}
M_speciesseason <- calcSplits(M = Mhat,
  split_CO = "Species",  data_CO = data_CO, split_SS = "Season", data_SS = data_SS)
summary(M_speciesseason)
plot(M_speciesseason)
```

There are so many vertical panels that it is difficult to glean any useful
information out of the graph. However, the panels may be transposed and graphed
for better interpretability:

```{r Transposed Species Season, fig.height = 7, fig.width = 7, fig.align = 'center'}
plot(transposeSplits(M_speciesseason))
```