---
title: "FlyingR: Predicting birds' flight range. A Tutorial."
author: "Brian Masinde"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{documentation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
citation_package: natbib
bibliography: references.bib
---

# Introduction
In this tutorial, a sample of 28 birds from the program Flight is used to 
demonstrate how to use the package to estimate flight ranges of migrating birds.
Note that in package `FlyingR` two approaches are implemented for this estimation.
The first from @penny75 and the second from @penny03 and @penny08. In this
tutorial focus is on the second approach (a time-marching approach).

## Basic usage with default settings.

```{r}
data("birds", package = "FlyingR")
results <-
  FlyingR::migrate(data = birds,
                  method = "cmm",
                  speed_control = 1,
                  min_energy_protein = 0.05)

# extract range as a vector
results$range
```


In this case, the muscle mass adjustment criteria is the constant muscle mass 
("cmm", meaning protein is not consumed from muscle mass). Run by holding 
true airspeed at start of flight constant, the alternative is maintain the 
ratio between true airspeed and minimum power speed constant (`speed_control = 0`). In this simulation, additional protein is consumed (5\% of the total energy)
and this comes from the airframe mass.

The default settings are as follows:

\begin{table}[H]
  \caption{Constants, denotation and default values}
  \label{constants}
  \begin{tabular}{|l|l|l|}
  \toprule
  \multicolumn{1}{|c|}{\textbf{Constant}}     & \textbf{denoted as} & \textbf{default value} \\ \midrule
  Profile power constant             & ppc        & 8.40          \\ \midrule
  Fat energy density                 & fed        & 3.9E+07       \\ \midrule
  gravity                            & g          & 9.81          \\ \midrule
  Mechanical conversion efficiency   & mce        & 0.23          \\ \midrule
  Induced power factor               & ipf        & 1.20          \\ \midrule
  Ventilation and circulation power  & vcp        & 1.10          \\ \midrule
  Air density                        & airDensity & 1.00          \\ \midrule
  Protein energy density             & ped        & 1.8E+07       \\ \midrule
  Body drag coefficient              & bdc        & 0.10          \\ \midrule
  %Basal metabolic rate               &            &               \\ \hline
  Mitochondria inverse power density & mipd       & 1.2E-06       \\ \midrule
  Muscle density                     & muscleDensity & 1060       \\ \midrule
  Protein hydration ratio            & phr           & 2.2        \\ \midrule
  Airspeed ratio                     & speedRatio    & 1.2        \\ \bottomrule
  \end{tabular}
\end{table}

These settings could be adjusted. For example, the default value for the induced
power factor is high, the recommended being 0.9. This can be adjusted as follows
`settings = list(ipf =0.9)`. Basic knowledge of defining lists is required.

```{r}
results <-
  FlyingR::migrate(data = birds,
                  method = "cmm",
                  settings = list(ipf = 0.9),
                  speed_control = 1,
                  min_energy_protein = 0.05)
```


Other methods are the constant specific work (`method = "csw"`) and constant 
specific power (`method = "csp"`).

```{r}
results <-
  FlyingR::migrate(data = birds,
                  method = "csw",
                  settings = list(ipf = 0.9),
                  speed_control = 1,
                  min_energy_protein = 0.05)


# obtain remaining body mass
results$bodyMass

# starting minimum power speed
results$startMinSpeed

# end of flight minimum power speed
results$endMinSpeed
```

\newpage 


# References
