---
title: "C++ binding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{C++ binding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package contains two [Rcpp](https://CRAN.R-project.org/package=Rcpp) functions,
which bind directly to the C++ implementation of the flexible polyline encoding.
These functions target a tight interface from C++ to R and reflect the arguments
and return values of their counterparts (`hf::encode_polyline` and
`hf::decode_polyline`) in the C++ implementation.

## Encode
Encoding a line is straight forward by passing a matrix (column order is
longitude, latitude) with its coordinates to `encode()`. The default precision
value is set to `5`. If a matrix with two columns is provided, the third
dimension is automatically set to `"ABSENT"`:

```{r encode2d}
library(flexpolyline)

line2d <- matrix(
  c(8.69821, 50.10228,
    8.69567, 50.10201,
    8.69150, 50.10063,
    8.68752, 50.09878),
  ncol = 2, byrow = TRUE
)

(encodes2d <- encode(line2d, precision = 5))
```

If a matrix with three columns is passed, a line with a third dimension is encoded.
The type of the third dimension can be passed to the encoding and supports the
following selection, that are specified via an integer value via the `third_dim`
argument:

* 0: ABSENT
* 1: LEVEL
* 2: ALTITUDE
* 3: ELEVATION (default)
* 4: RESERVED1 (not available)
* 5: RESERVED2 (not available)
* 6: CUSTOM1
* 7: CUSTOM2

By default the precision of the third dimension `third_dim_precision` is set to
the same value as the precision value for the longitude and latitude dimensions.
```{r encode3d}
line3d <- matrix(
  c(8.69821, 50.10228, 10.11111,
    8.69567, 50.10201, 20.22222,
    8.69150, 50.10063, 30.33333,
    8.68752, 50.09878, 40.44444),
  ncol = 3, byrow = TRUE
)

(encodes3d <- encode(line3d, precision = 5, third_dim = 3, third_dim_precision = 5))
```


## Decode

In order to decode the lines from above, the encoded line string is passed to
the `decode()` function. The function detects from the encoded string if it is a
two-dimensional or three-dimensional line:
```{r decode2d}
decode(encodes2d)
```

In the case of a three-dimensional line, the third column is named after the name
of the third dimension that was chosen during the encoding (in this example
`"ELEVATION"`):
```{r decode3d}
decode(encodes3d)
```


## References

* [Flexible Polyline Encoding by HERE](https://github.com/heremaps/flexible-polyline)
* [Encoded Polyline Algorithm Format](https://developers.google.com/maps/documentation/utilities/polylinealgorithm)
* Inspired by the [googlePolylines](https://github.com/SymbolixAU/googlePolylines) package
