---
title: "2. The Copula Recursive Tree"
bibliography: ../inst/REFERENCES.bib
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{2. The Copula Recursive Tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

```{r setup, include = FALSE}
library(cort)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

In this vignette, we will use the Cort algorithm to fit a copula on a given simulated dataset based on clayton copula simulations. We will show how the algorithm can be used to produce a bona-fide copula, and describe some of the parameters. 

# Dataset

First, let's create and plot the dataset we will work with. For that, we'll use the gamma frailty model for the Clayton copula (but it'll work for any other completely monotonous archimedean generator), as it is done in the `copula` package, see [there](https://github.com/cran/copula/blob/6e0a5f0332776c24b2f53a81b11a8c66cc3f996d/R/claytonCopula.R#L82). The following code is directly taken from the previous link, from the `copula` package : 

```{r clayton-frailty-model}
psi <- function(t,alpha) (1 + sign(alpha)*t) ^ (-1/alpha) # generator
rClayton <- function(n,dim,alpha){
  val <- matrix(runif(n * dim), nrow = n)
  gam <- rgamma(n, shape = 1/alpha, rate = 1)
  gam <- matrix(gam, nrow = n, ncol = dim)
  psi(- log(val) / gam,alpha)
}
```


The following code simulates a dataset and then visualise it : 

```{r dataset} 
set.seed(12)
n = 200 # taken small to reduce runtime of the vignette.
d = 4
n_trees = 5 # taken small to reduce runtime of the vignette.
number_max_dim_forest = 2 # taken small to reduce runtime of the vignette.

data <- matrix(nrow=n,ncol=d)
data[,c(1,4,3)] = rClayton(n=n,dim=d-1,alpha=7)
data[,2] = runif(n)
data[,3] <- 1 - data[,3]

pairs(data,cex=0.6)
  
```

We can clearly see that the second marginal is independent form the rest. In the following we will use this package to fit this dependence structure. 


# Fitting the Cort copula

Now that we have a dataset, we can run the Cort algorithm on it. In the implementation proposed here, this is done via the `cort::Cort()` function, passing first the dataset, and then various parameters. See `?Cort` for a detailed list of parameters. Note that the verbosity level is quite progressive: We will here put it on 4 to see the splitting decisions that the algorithm is making.

```{r run_cort} 
set.seed(12)
(model = Cort(data,verbose_lvl = 1))
```

Looking at the top of the output, we see that the first thing the algorithm did was removing the second dimension due to the independence test. Now that the copula is fitted, we have access to numerous of it's methods. Two plotting functions are exported with this model, the `pairs` function is implemented at a very low level in the class hierarchy and hence is working with almost all copulas of this package, but the `plot` function is only implemented for Cort.  


```{r,fig.cap="Pairs-plot of original data (in black, bottom-left corner) versus a simulation from the model (in red, top-right corner)"}
pairs(model)
```

```{r,fig.cap="Gray boxes representing 2-d projections of the fitted density. In red, the imputed data points."}
plot(model)
```


We see that there are some noise with point were there should not be. A bagged version of the model is accessible via the `CortForest` class, and might be able to correct these problems.






